### 2016-07-01 configuration_files.tex
New pieces of the structure

\label{sec:testbed}

% We proceed to define the testbed environment by each of its conforming
% parts. We later indicate a procedure to set up and configure the testbed
% for several \ac{Raspi}s. In this procedure, we summarize all the key details
% regarding on setting the system files, configuration packages, tools and connectivity
% for the \ac{Raspi}s in a centralized fashion.

%Afterwards, we indicate how to cross-compile C++ code and the Kodo
%network coding library in an easy way.
%Finally, we provide further information about Kodo itself in terms
%of testing, other platforms supported and source code documentation
%for further references.

%\subsection{System Overview and Design Criteria}

A sketch of the testbed is depicted in Fig.~\ref{fig:testbed_setup}.
The testbed consists of up to 100 \ac{Raspi}s of different models.
More specifically, in our design we consider: Raspberry Pi 1 model B rev.~2,
Raspberry Pi 2 model B V1.1 and Raspberry Pi 3 model B V1.2.
All \ac{Raspi}s are each equipped with a memory card, a wired and wireless
network interface and a power supply. All the \ac{Raspi} are connected to a common
\ac{LAN} that provides internal and external connectivity. Without loss of generality,
in our case they are connected to the university network using their wired Ethernet interface
that is named \texttt{eth0} according to the legacy naming convention of Ethernet
interfaces in Linux~\cite{PredictableNetworkInterfaceNames}. We consider the
university network since we aim the testbed to be used by students and professors
to perform measurements and experimentation of controlled and reproducible scenarios
as part of academic research. We intend to configure all \ac{Raspi}s identically,
while still allowing the end-users to alter how each \ac{Raspi} behaves and is configured.

\begin{figure}[ht!]
\centering
\includegraphics[width=0.45\textwidth]{images/testbed_setup3.eps}
\caption{Testbed setup}
\label{fig:testbed_setup}
\end{figure}

%This significantly eases the management of the \ac{Raspi}s
%as \ac{SSH} eliminates the need for monitors and keyboards for the \ac{Raspi}s.

In Fig.~\ref{fig:testbed_setup} and what follows, we will refer
as the testbed administrator to the person(s) in charge of setting up and configuring
the testbed with administrator privileges from the \ac{OS} point of view.
The setting and configuration procedure commands are performed the testbed administrator in a
regular \ac{PC} running a Linux distribution. Although in principle the administrator
Linux distribution is not a restriction, we present our procedure in a Debian-based Linux
distribution. Finally, our design considers to fetch the \ac{Raspi}s
root filesystem from a \ac{HTTP} server. This enables the testbed administrator to
make filesystem in one location. The \ac{Raspi}s will then get the changes after a reboot.
All \ac{Raspi}s will be configured to run a \ac{SSH} daemon for easy remote access
within the university network. We requested the university \ac{IT} department to
configure the university \ac{DHCP} server to assign each \ac{Raspi} a static
\ac{IP}. This eliminates the requirements of monitors and keyboards with the
\ac{Raspi}s for non-graphical applications.

Based on a regular Raspbian Linux image, we explain how
to customize it to include desired packages for performing routinary tasks and set it
in all the \ac{Raspi}s. Also, when working on a testbed with many devices, it may not always be
desired to make persistent changes to the devices. Thus, we later present how to utilize
stacked filesystems to enable both persistent and temporary
storage. This step of the procedure will be optional if the testbed
administrator decides to write the whole image into a memory card. \todot{Check the
veracity and correctness of the previous statement. Question: Shoudln't we use
an overlay FS regardless of how the memory card is configured?}.
Although being simpler than a stacked filesystem, writing the whole Linux
filesystem to several memory cards may complicate future
maintenance of the testbed since updates need to be performed in each \ac{Raspi}.
Even when tools such as Python Fabric may ease this task, the maintenance might still
be cumbersome in some cases. Therefore, we propose and encourage the alternative centralized
approach where the initial \ac{RAM} filesystem is customized in a way that
enables \ac{Raspi}s to fetch the its root filesystem from a \ac{HTTP} server as
mentioned previously. The following procedure explains how the testbed may be setup and configured.

%\textcolor{red}{The testbed should be available for students to perform simulations and
%measurements as part of their semester projects. We intend to configure
%all \ac{Raspi}s identically, while still allowing students to alter
%how each \ac{Raspi} behaves and is configured.
%\\\\
%When working on the testbed with this many devices, it may not always be
%desired to make persistent changes to devices. We present how the concept
%of stacked filesystems may be used to enable the means of both persistent
%and temporary storage.}

%The testbed should be available for students to perform simulations and
%measurements as part of their semester projects. The \ac{Raspi}s should
%be configured identically, but allows the students to alter the
%configurations and how each \ac{Raspi} behaves.
%
%When working on a testbed with many devices, it may not always desired
%to make persistent changes to devices. We present how the concept of stacked
%filesystems may be used to enable the means of both persistent and
%temporary storage.

%A \ac{HTTP} server will be used to store Linux configuration files as well
%as the root filesystem in the centralized approach.

%The following procedure will explain how the testbed may be setup and
%configured both in the traditional distributed fasion where the
%entire Linux filesystem is written to an memory card, but also in a more
%centralized fasion where 
%is written to
%Based on the popular Debian based Raspbian Linux distribution
%We
%first present how a standard Linux distribution for the \ac{Raspi}s
%may be customized

\subsubsection{Installing Raspbian}

To get started, we first need to install an \ac{OS} on the
\ac{Raspi}s. We will use the popular Debian-based Raspbian
Linux~\cite{raspbian}

We will be using a \ac{PC} running a Debian-based Linux distribution
to setup a customized Raspbian Linux image that can be written to
a memory card and work in any \ac{Raspi} version 1 or 2.

Raspbian is made available in two bundles: Raspbian and Raspbian lite.
The difference between the two is that Raspbian contains a pre-installed desktop
environment, and Raspbian lite is by default only accessible through a 
command shell. Both Raspbian and Raspbian lite will both by default be remotely
accessible using \ac{SSH}. \ac{SSH} provides the means to remotely
connect to another machine through a command shell.
%In fact it is the successor
%of \ac{Telnet} that has been designed with network security in mind.

The \ac{Raspi}s in our testbed are not connected to monitors, so we
have decided to go with Raspbian lite.
%Our testbed does not have monitors so we will install Raspbian lite.
A desktop environment can be installed using the package manager
(apt-get) in case a desktop environment will be desired at a later
time.

Below is the procedure we followed to setup the \ac{Raspi}s in our testbed. We
have performed a simple setup where each \ac{Raspi} is equipped with a
memory card that contains a slightly customized Raspbian lite image. The image
of the \ac{OS} in all \ac{Raspi}s will be identical.

The overall procedure to customize the official Raspbian lite image
are:
\begin{enumerate}
    \item Download Raspbian lite
    \item Alter Raspbian lite. e.g. browse, modify, add, and delete files
        in the official Raspbian lite image
    \item Change root, i.e. change root filesystem into the Raspbian lite
        image to update and install software packages
    \item Write image to memory cards
\end{enumerate}

Hashmark (\#) and dollar (\$) will be used throughout the code blocks in
the paper to indicate if an operation needs to be run with root permissions or
user permissions.

\subsection{Download Raspbian lite}

The latest Raspbian lite bundle can be downloaded from Raspbian's official
webpage~\cite{raspbian}.
%To download the latest Raspbian lite, we go to the url
%\url{http://downloads.raspberrypi.org/raspbian_lite/images/}.
At the time of this writing, the latest available bundle was
\texttt{2016-05-27-raspbian-jessie-lite.zip}.

To ensure that the content of the bundle does not change, this procedure
will be based on that particular version of Raspbian lite which we have
made available at~\cite{tunescode_webpage}. All other files used in this paper
will also be available there unless specified otherwise.

%We want to proceed downloading the bundle in a shell instead of the browser.
%To do this, first open a command shell on a linux machine. This guide is
%assuming a Debian-based distribution, but there should only be minor differences
%if you have another package manager. i.e. the package managers and name of software packages.

To get started, open a Linux shell (terminal) and declare the variables below.
The \texttt{URL} and \texttt{IMAGE} specifies where the Linux bundle will be
located and \texttt{WORKDIR} specifies a working directory where we will 
store Raspbian and customize it.
Do not type in \$ and \# in the shell. Those indicates if a command can be
executed as a regular user or if root permissions are required.
Running a command as root can on most systems be done by putting
\texttt{sudo} in front of the command. This is illustrated in the
following code block 
with the command \texttt{whoami} that prints the username. Lines within a code
block without leading \$ or \# is terminal output or content of a file.

%After the terminal has been opened, we start by declaring a few variables to reduce
%repeated typing. The first variables we declare is the Raspbian image name.
%Notice that the extension has been omitted. This is because the image has been
%packed into a zip file. The other variable we declare is a working directory.
%This is where we will download the image to and work on it. In other words, it
%will be stored in \texttt{/home/<username>/Raspbian}

%To download Raspbian lite, we go to the \url{http://downloads.raspberrypi.org}.
%There should be a folder called raspbian\lite and
%to find the download page at which Raspbian lite should be located. Instead of
%downloading with the browser, we just extract the download url to the latest
%image. At the time of this writing, that was

% Define variable]
\begin{lstlisting}[]
$ export URL="http://kom.aau.dk/project/TuneSCode/raspi/"
$ export IMAGE="2016-05-27-raspbian-jessie-lite"
$ export WORKDIR="${HOME}/Raspbian"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%A final thing to notice is the way \$ and \# are used in the code blocks. \$
%means that the command can be run with normal user permissions where \# indicates that the
%command needs root permissions.
%Running a command with root permissions can on most systems be done by putting
%the command "sudo" in front of each line. This is illustrated in the code below
%with the command "whoami" that prints the username of the caller. Lines without
%leading \$ or \# is the terminal output.

% Root example
\begin{lstlisting}[]
$ whoami
<USERNAME>
$ sudo whoami
root
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Next, make the working directory and used \texttt{cd} to change into the working
directory. Then \texttt{wget} can be used to download Raspbian lite before
unpacking the \texttt{zip} file:

% Download and unpack image
\begin{lstlisting}[]
$ mkdir -p ${WORKDIR}
$ cd ${WORKDIR}
$ wget ${URL%/*}/${IMAGE}.zip
$ unzip ${IMAGE}.zip
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%A more recent version of Raspbian lite may be available at~\cite{raspbian}.
%\url{http://downloads.raspberrypi.org/raspbian_lite/images}.

\subsection{The image}

After Raspbian lite has been unpacked, there should be an \texttt{img} file in the directory.
\texttt{fdisk} can be used to display the content of the image. Parse the arguments
"-u sectors" to display the sizes in sectors and "-l" to display the partitions within the
image and and automatically exit \texttt{fdisk} again:

% Check out the image   % The dollar hack was to fix vim syntax
\begin{lstlisting}[literate={DOLLAR}{\$}1]
DOLLAR fdisk -u sectors -l ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 1.3 GiB, 1387266048 bytes, 2709504 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

Device                               Boot  Start     End Sectors  Size Id Type
2016-05-27-raspbian-jessie-lite.img1        8192  137215  129024   63M  c W95 FAT32 (LBA)
2016-05-27-raspbian-jessie-lite.img2      137216 2709503 2572288  1.2G 83 Linux
\end{lstlisting}
\FloatBarrier

The output provides knowledge about the image. The image is in total
2709504 sectors (1.3GiB) in size and contains two
partitions. The first partition starts at sector 8192 and
the other partition starts at sector 137216. The partitions is of type FAT32
and Linux and are 63M and 1.2G big. This indicates that the first partition
is most likely a boot partition and the second partition is a traditional Linux file
system. Actually, it is the Linux root filesystem (i.e. /).

\subsection{Resize image}
1.2G is not very much to store the root filesystem given that it is desired to
alter the files and install additional packages. It may therefore be
necessary to increase the partition size.
The following illustrates how the image and its root filesystem
can be expanded by one \ac{GB}.

First, expand the image one \ac{GB} and wait for the data to be completely written
to the disk:
\begin{lstlisting}[]
$ dd if=/dev/zero bs=1M count=1024 >> ${IMAGE}.img && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Use \texttt{fdisk} to see that the image is now one \ac{GB} larger:
\begin{lstlisting}[literate={DOLLAR}{\$}1]
DOLLAR fdisk -u sectors -l ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 2.3 GiB, 2461007872 bytes, 4806656 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

Device                               Boot  Start     End Sectors  Size Id Type
2016-05-27-raspbian-jessie-lite.img1        8192  137215  129024   63M  c W95 FAT32 (LBA)
2016-05-27-raspbian-jessie-lite.img2      137216 2709503 2572288  1.2G 83 Linux
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, to expand the root filesystem it is required to first remove the Linux partition
and then add it again one \ac{GB} bigger.
Use \texttt{fdisk} to alter the partition table. The \texttt{echo}'s within the parenthesis
will be interpreted as
key-presses in \texttt{fdisk} and hence delete the last partition, create a new
primary partition, print new partition table, and write new partition table
to the image:

\begin{lstlisting}[]
$ (echo d; echo ; echo n; echo p; echo ; echo 137216; echo ; echo p; echo w) | fdisk ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Use a loopback device to make the Raspbian image available as a block device: 
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\texttt{lsblk} may be used to list block devices. It appears that the image is
available as \texttt{/dev/loop0} which has two partitions \texttt{loop0p1} and
\texttt{loop0p2}:
%Use \texttt{lsblk} to view the partitions:

\begin{lstlisting}[]
# lsblk
NAME      MAJ:MIN RM  SIZE RO TYPE MOUNTPOINT
...
loop0       7:0    0  2.3G  0 loop 
|-loop0p1 259:2    0   63M  0 loop 
|-loop0p2 259:3    0  2.2G  0 loop
...
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Check and resize the root partition:
\begin{lstlisting}[]
# e2fsck -f ${DEV}p2
# resize2fs ${DEV}p2
\end{lstlisting}
\FloatBarrier

%We see that the boot and root partitions starts at sector 8192 and 137216 respectivly.
%We also notice that each sector is 512 bytes.
%On newer systems, we can mount the device easier using "losetup --show -f -P IMAGE"
%Mount root partition:

\subsection{Mount image}

Browsing and altering the files in the image is easy. Simply mount the partitions.
%To browser and alter the files in the image, it is possible to mount the partitions.

Lets first mount the root partition. This is done by creating an empty directory
that we will be used as a mountpoint. We name it \texttt{root} and create it in the
working directory before mounting the root filesystem onto the mountpoint.
%After this, mount the partitions:
%After this,
%we mount the partition starting at sector 137216. To provide this offset information
%to mount, we need to find the offset in bytes. From the fdisk command we could also
%see that each sector is 512 bytes. Thus, we multiply the sector size and the sector
%start to get the offset in bytes.

\begin{lstlisting}[]
$ export ROOTDIR="${WORKDIR}/root"
$ mkdir -p ${ROOTDIR}
# mount ${DEV}p2 ${ROOTDIR}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((137216*512)) ${IMAGE}.img ${ROOTDIR}

The root filesystem already has a boot directory that can be used as
the mountpoint for the boot partition.
%After this, we can also mount the boot partition inside the newly mounted
%root filesystem.
This is convenient because that partition will be mounted
on this directory when a \ac{Raspi} starts up with a memory card containing
the image. %It is therefore the natural place to mount it for later purposes.
Now, mount boot partition:
\begin{lstlisting}[]
# mount ${DEV}p1 ${ROOTDIR}/boot
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((8192*512)) ${IMAGE}.img ${ROOTDIR}/boot

%We can now change all the files in the disk image as desired.
It is now possible to change all files within the Raspbian image as desired.
We will take advantage of this to edit configuration files, append new files,
and even update and install packages.

@@

\subsection{Installation an overlaying filesystem}

Our goal was to enable the users of the testbed to change however the
\ac{Raspi} are configured while still being able to revert those changes easily.

We do this by storing a users work in the \ac{Raspi}s in \ac{RAM}. Thus a
\ac{Raspi}, when a \ac{Raspi} is rebooted, it will forget all changes.

A way to do this is using stacked filesystem. Stacking filesystem enables
a filesystem to overlay another filesystem. Whenever a file is accessed,
the upper filesystem will forward the request to the lower filesystem in
case it does not have it itself. If it has it, it will simpy return the file.

This idea can be used in our setup to mount the root filesystem (i.e. Raspbian
lite) in the \ac{Raspi}s duing startup as read-only. Thus, the files will not
be changable. Some programs needs to change files and the testbed admin might
want to change configurations. To enable this, it is possible to
overlay a filesystem that is mounted in \ac{RAM} as read/writable on top of
the regular root. Thus, reading a file may return a file from the lower filesytem,
but if it is stored, it will be saved in the upper filesystem. Accessing this file
again will return the stored file from the upper layer.

%We do this using an overlay filesystem. The idea is to let the \ac{Raspi}s
%mount their root filesystem as 
%
%We do this using an overlay filesystem (overlayfs).
%Overlayfs enables a filesystem to be stacked on top of another filesystem. Thus,
%if the \ac{Raspi}s mount their root filesystem (/) as read-onlyif we mount our root (/) filesystem as 
%
%
%Overlayfs enables filesystems
%to be stacked on top of each other. We will be using this to make a lower
%filesystem and an upper filesystem. The lower filesystem will be our
%traditional filesystem and the upper filesystem will be mounted in tmpfs
%(i.e. mounted in \ac{RAM}).
%
%Thus, when a file is requsted from stacked filesystem, it will first attempt
%to fetch the file in the upper filesystem if it does not exists there, it will
%request it from the lower filessytem.

Assuming that you are still \texttt{chrooted} into the Raspbian lite filesystem, we
can setup the overlay filesystem (overlayfs).

There already exists implementations overlaying the root filesystem. We will
use an existing implementation
available at github: \cite{overlayroot}.

We installed git in a previous step, so lets clone the repository. The below
code stores it in /tmp which is really mounted in \ac{RAM}. Thus, the files
will disappear when the machine is rebooted.

% Get files
\begin{lstlisting}[]
(chroot) # OVERLAYROOTDIR="/tmp/overlayroot"
(chroot) # git clone https://github.com/chesty/overlayroot.git $OVERLAYROOTDIR
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

To enable the overlaying filesystem it is required to generate an initial \ac{RAM}
file system (initramfs). Initramfs is an initial filesystem that gets loaded into
\ac{RAM} during the startup process of a linux machine to prepare the real
filesystem.

For this purpose, we need BuxyBox:

% Install required packages
\begin{lstlisting}[]
(chroot) # apt-get install busybox
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Add and activate the overlaying filesystem, we need to add the scripts and tell initramfs
to load the overlaying filesytem:

% Setup initramfs
\begin{lstlisting}[]
(chroot) # cp ${OVERLAYROOTDIR}/hooks-overlay /etc/initramfs-tools/hooks/
(chroot) # cp ${OVERLAYROOTDIR}/init-bottom-overlay /etc/initramfs-tools/scripts/init-bottom/
(chroot) # echo "overlay" > /etc/initramfs-tools/modules
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%Finally, lets generate the initial \ac{RAM} filesystem. This is done using the command
%\texttt{mkinitramfs}, that will by default look for the kernel modules of the current system. Remember we are
%chrooted, so we need to specify the correct kernel modules to look for. These
%can be found in "/lib/modules".
Now, the initial \ac{RAM} filesystem can be generated. \ac{Raspi} version 1
needs a different kernel than \ac{Raspi} version 2 and version 3. Thus, start
by finding the available kernel modules:
%Because \texttt{initramfs} is executed within a
%\texttt{chrooted} environment, it will not be able to detect the
%kernel modules in the \texttt{chrooted} environment. The kernel modules will

% Create an initramfs
\begin{lstlisting}[]
(chroot) # ls /lib/modules/
4.4.13+  4.4.13-v7+
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Kernel "4.4.13+" is for \ac{Raspi} version 1 and kernel "4.4.13-v7+" works
for \ac{Raspi} version 2 and version 3.

Lets proceed by generating an initial \ac{RAM} filesystem for both kernels:

%Lets proceed with kernel modules "4.4.13+".
%Lets create an initial ram filesystem for both kernels.

%Thus, to run mkinitramfs, type in the correct kernel version:

% Create an initramfs
\begin{lstlisting}[]
(chroot) # mkinitramfs -o /boot/init.gz -k 4.4.13+
(chroot) # mkinitramfs -o /boot/init-v7.gz -k 4.4.13-v7+
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%IT SEEM THAT THE RASPIS DEMANDS DIFFERENT KERNEL MODULES AFTER ALL.
%RASP-VERSION1=4.4.13+, RASP-VERSION2=4.4.13-V7+, RASP-VERSION3=4.4.13-V7+

Although there might be printet a few warnings, the above commands should successfully
generate working initial \ac{RAM} filesystems.

An initial \ac{RAM} filesystem needs to be called by the bootloader. In Raspbian
this is done by adding a command to \texttt{config.txt} in the boot partition.

If the system should be run in a \ac{Raspi} version 1, then use \texttt{init.gz}:
%The bootloader should be made aware of the initramfs that we will generate:
\begin{lstlisting}[]
(chroot) # echo "initramfs init.gz" >> /boot/config.txt     # Only for Raspberry Pi version 1
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

If the system should be run in a \ac{Raspi} version 2 or 3, then use \texttt{init-v7.gz}:
\begin{lstlisting}[]
(chroot) # echo "initramfs init-v7.gz" >> /boot/config.txt  # Only for Raspberry Pi version 2 and 3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

At this point, it is no longer required to be \texttt{chrooted} and the image
can be written to a memory card. The following commands exits the \texttt{chrooted}
environment, unmount all partitions, and detaches the loopback device:

% Exit chroot, umount, and write to memory card
\begin{lstlisting}[]
(chroot) # exit
# cd ..
# umount --recursive ${ROOTDIR}
# losetup -d ${DEV}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\subsection{(Optional) Persistent subdirectories for distributed Linux image}

EMPHASIZE THAT THIS SETUP IS ONLY FOR THE SETUP WHERE THE ROOT IMAGE ARE
WRITTEN TO THE MEMORY CARDS

Provided the stacked filesystem was configured, it may be desired to have
some directories where files are not flushed upon rebooting the \ac{Raspi}s.
The following procedure will create an extra partition in the image for
the \ac{Raspi} users home directory that will be made storage persistent.
%Provided that a stacked filesystem was configured to make the root
%filesystem forget chances upon reboots is not always desired.

Expand image according to the desired home directory size. Keep in mind not
to make the image bigger than the target memory card.
% Expand image
\begin{lstlisting}[]
$ dd if=/dev/zero bs=1M count=1024 >> ${IMAGE}.img && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create home partition after the root partition. Use \texttt{fdisk} to
find the next available sector in the image. In our case, the new
partition should start at sector 4806656:
\begin{lstlisting}[]
$ fdisk -lu ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 3.3 GiB, 3534749696 bytes, 6903808 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

Device                               Boot  Start     End Sectors  Size Id Type
2016-05-27-raspbian-jessie-lite.img1        8192  137215  129024   63M  c W95 FAT32 (LBA)
2016-05-27-raspbian-jessie-lite.img2      137216 4806655 4669440  2.2G 83 Linux
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create new partition:

\begin{lstlisting}[]
$ (echo n; echo p; echo ; echo 4806656; echo ; echo p; echo w) | fdisk ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create loopback device: 
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Format new partition:
\begin{lstlisting}[]
# mkfs.ext4 ${DEV}p3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The last thing to remember is to inform Raspbian to mount the home partition
we created during startup. This can be done by adding an entry in \texttt{fstab}:
\begin{lstlisting}[]
# mount ${DEV}p2 ${ROOTDIR}
# sudo sh -c "echo '/dev/mmcblk0p3 /home ext4 defaults,noatime 0 2' >> ${ROOTDIR}/etc/fstab"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The home folder was originally located in the root filesystem. We should
therefore copy the content to the new home partition:
\begin{lstlisting}[]
# mount ${DEV}p3 ${ROOTDIR}/mnt
# mv ${ROOTDIR}/home/* ${ROOTDIR}/mnt/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

That is it. unmount partitions and detach loop device:
\begin{lstlisting}[]
# umount --recursive ${ROOTDIR}
# losetup -d ${DEV}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\subsection{Write image to memory card}

EMPHASIZE THAT THIS SETUP IS ONLY FOR THE SETUP WHERE THE ROOT IMAGE ARE
WRITTEN TO THE MEMORY CARDS

The final step is to write the image to all the memory cards before they can
be used in the \ac{Raspi}s.

%This is done by first leaving the \texttt{chrooted}
%filesystem and unmounting all the directories we mounted:

% Exit chroot, umount, and write to memory card
%\begin{lstlisting}[]
%# exit
%# cd ..
%# umount --recursive ${ROOTDIR}
%# losetup -d ${DEV}.img
%\end{lstlisting}
%\FloatBarrier
%\vspace{-5mm}

%If successfully unmounted, you can 
Insert a memory card in your machine. It will be available as
\texttt{/dev/mmcblkX} or \texttt{/dev/sdX}

%In most systems, this memory card was available under \texttt{/dev/mmcblk0}, but this may be
%different on different machines. For internal memory card readers, it will
%probably be named /dev/mmcblkX where X is some number and if the memory card
%reader is connected through \ac{USB}, then it is likely called /dev/sdX.
Make absolutely sure to write to the correct device as everything will be 
overwritten!
Inserting/Ejecting the memory card a couple of times and using the commands
below may help deducing the correct device:
%The commands below may help deduce which device is the correct:
%Below is a few commands that may help you to deduce which device is correct:

\begin{lstlisting}[]
# lsblk
# df -h
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

For us, the device was \texttt{/dev/mmcblk0}. Write the image to a memory card:
\begin{lstlisting}[]
# dd if=${IMAGE}.img of=/dev/mmcblk0 bs=4M && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The memory card can now be started in a \ac{Raspi}.

>>>>

\label{sec:configuration_files}

@@
